---
title: "ipmr Introductory Tutorial"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    highlight: pygments
runtime: shiny_prerendered
tutorial:
  version: 0.0.0.3
---


```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


## Overview

Welcome to the `ipmr` pre-workshop tutorial! This tutorial will teach you the basics of how to use `ipmr` to implement Integral Projection Models (IPMs), as well as how to work with model objects to complete further analyses. We will use real data to investigate optimal management strategies for an invasive population of *Carpobrotus* species (iceplant) in Israel. We will also walk through a few additional analyses, and how to use helper functions from `ipmr` to compute them. 

This tutorial contains demonstrations of some newer features in `ipmr` available in version `0.0.3`. This is now on CRAN as of mid-July 2021. Please update to this version if you have not already, as some of the functionality will not be available otherwise.

```{r eval = FALSE, echo = TRUE}

install.packages("ipmr")

```

### Finding help

**Package website**

The project website is https://levisc8.github.io/ipmr. This website contains all of the help files for the package functions, as well as a number of longer articles on how to implement more complicated models, including two kinds of stochastic IPMs, general IPMs (_i.e._ IPMs with a mix of discrete and continuous state variables), and IPMs with mixtures of discrete and continuous environmental variation.

**Bug reports**

Please use the issue tracker here: https://github.com/levisc8/ipmr/issues. If you are not comfortable with GitHub, then you can also email Sam: <levisc8@gmail.com>.

**Instructor Email Addresses**

Sam: <levisc8@gmail.com>/<sam.levin@idiv.de>

Sanne: <sanne.evers@idiv.de>

### Notation for the tutorial

This exercises in this tutorial use the mathematical notation of IPMs to demonstrate how `ipmr` works with these models. This slide is a quick guide for the notation convention

#### Traits, kernels and functions

  - Population trait distributions, such as size, weight, height, or color will be denoted $n(z,t)$ and $n(z',t+1)$. These are the trait distributions for trait $z$ at time $t$ and $z'$ at time $t+1$, respectively.
  
  - Bivariate kernels (*i.e.* kernels that are functions of $z$ and $z'$) will be capitalized and have at least two arguments (*e.g.* $P(z',z)$, $K(z',z)$). 
  
  - Univariate functions (*i.e.* a regression model `survival ~ size`) will be lower case and have 1 argument (*e.g.* $Logit(s(z))$, or $Log(r_n(z))$). 
  
  - Constant demographic parameters will be lower case and not be a function of anything (*e.g.* $r_g$).

## Iceplant IPM

```{r, child = 'sub-pages/iceplant_study.rmd'}

```



